---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# multipleOutcomes

<!-- badges: start -->
[![Project Status: Active – The project has reached a stable, usable state and is being actively developed.](https://www.repostatus.org/badges/latest/active.svg)](https://www.repostatus.org/#active)
<!-- badges: end -->

The goal of multipleOutcomes is to fit statistical models for multiple outcomes simultaneously. It computes estimates of parameters across fitted models and returns the matrix of asymptotic covariance. Various applications of this package, including PATED (Prognostic Variables Assisted Treatment Effect Detection), multiple comparison adjustment, conditional power are illustrated.


## Installation

You can install the development version of multipleOutcomes from [GitHub](https://github.com/) with:

``` r
# install.packages("pak")
pak::pak("zhangh12/multipleOutcomes")
```

## Example

Let's get started by analyzing a really randomized trial data

```{r warning = FALSE, message = FALSE}
library(multipleOutcomes)
library(dplyr)

options(digits = 2)

data(indo)

fit <- pated(
  outcome ~ rx, 
  
  risk ~ rx, 
  gender ~ rx, 
  sod ~ rx, 
  pep ~ rx,
  recpanc ~ rx,
  psphinc ~ rx, 
  precut ~ rx, 
  difcan ~ rx,  
  amp ~ rx, 
  paninj ~ rx, 
  acinar ~ rx,
  asa81 ~ rx, 
  asa ~ rx,
  prophystent ~ rx, 
  therastent ~ rx, 
  pdstent ~ rx, 
  
  data = indo,
  family = c('binomial', 
             rep('gaussian', 1), 
             rep('binomial', 15))
)

plot(fit)
```

For each of the covariates, a regression model `covar ~ rx` is fitted to compute the slope coefficient $\hat b$. The correlation between $\hat b$ and $\hat\beta$, the unadjusted estimate of treatment effect (`outcome ~ rx`), are computed in the column `corr`. 

```{r}
print(fit)
```



