---
title: "PATED: Prognostic-Assisted Treatment Effect Detection in Randomized Trials"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{PATED: Prognostic-Assisted Treatment Effect Detection in Randomized Trials}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

## Overview

In a properly randomized clinical trial, baseline covariates are balanced
between treatment arms *in expectation*. PATED (Prognostic-Assisted
Treatment Effect Detection) exploits that fact: by jointly modelling the
primary endpoint together with a panel of prognostic baseline covariates
and accounting for their joint covariance, the estimated treatment effect
on the primary endpoint can be adjusted to remove chance imbalance,
shrinking its standard error without introducing bias.

`pated()` returns a data frame with one row per term. The first two
rows are the adjusted (`method = "PATED"`) and unadjusted
(`method = "Standard"`) estimates of the primary treatment effect,
each with its standard error and p-value. The remaining rows hold the
per-arm regression of each prognostic covariate on the treatment
indicator (`method = "Prognostic"`), again with estimates, standard
errors, p-values, and the estimated correlation with the primary
endpoint. The function does not report a Wald statistic or relative
efficiency directly; those are simple derived quantities and we
compute them inside this vignette via a small helper.

This vignette walks through eight publicly available randomized trial
datasets distributed in CRAN packages. Each example fits a Cox model for
the primary time-to-event endpoint together with GLMs for the baseline
covariates, all sharing the treatment indicator. Datasets that bundle
disparate cohorts or that originate from observational studies are
deliberately excluded so that the randomization assumption underlying
PATED is satisfied.

### Reading the diagnostic plot

Diagnosis is delivered through the `plot()` method, which converts
each row of the `pated` object into a Wald z-statistic
(`estimate / stderr`) and arranges them on a forest-style chart:

- The **red** point at the top is the PATED-adjusted z for the
  primary endpoint.
- The **blue** point just below is the *unadjusted* z for the same
  endpoint.
- The **gray** points are the per-arm z-statistics for each baseline
  covariate (`covariate ~ arm`), ordered by their estimated
  correlation with the primary outcome (strongest at the top).

Dashed vertical lines mark `z = 0` and `z = +/-1.96`. Two things to
look for:

1. **Baseline balance.** Under randomization the gray prognostic
   points should scatter around zero with only a few drifting past
   `+/-1.96` by chance. A systematic shift to one side, or many
   covariates outside the +/-1.96 band, suggests imbalance worth
   investigating before relying on the adjusted estimate.
2. **Power gain.** PATED removes the component of the primary effect
   that is explained by chance baseline imbalance. When the red PATED
   point sits noticeably further from zero than the blue unadjusted
   point, the prognostic panel has shaved variance off the primary
   estimate; when red and blue nearly coincide, the covariates added
   little.

### Relative efficiency helper

Relative efficiency (RE) of PATED over the unadjusted analysis is the
ratio of their sampling variances. We compute it from the two
treatment-effect rows of the `pated` return:

\[
\text{RE} = \left(\frac{\text{stderr}_\text{Standard}}{\text{stderr}_\text{PATED}}\right)^2
\]

A value larger than 1 means PATED produced a tighter standard error
than the unadjusted analysis on the same data.

```{r setup, message = FALSE, echo = TRUE}
library(multipleOutcomes)
library(dplyr)
library(survival)
options(digits = 2)

# pated() requires a subject identifier; add one if the data lack it.
add_pid <- function(df) {
  df$pid <- paste0("s-", seq_len(nrow(df)))
  df
}

# Derive RE from the two primary-effect rows that pated() returns.
relativeEfficiency <- function(obj){
  adj   <- obj[obj$method == 'PATED',    'stderr']
  unadj <- obj[obj$method == 'Standard', 'stderr']
  (unadj / adj)^2
}

printObject <- function(obj){
  message(gsub('_', '::', deparse(substitute(obj))))
  message(paste0('Relative Efficiency: ',
                 format(relativeEfficiency(obj), digits = 3)))
  print(obj)
}
```

## Smoking cessation: `asaur::pharmacoSmoking`

Steinberg et al. (2009) randomized 125 smokers (61 combination,
64 patch-only) to a triple-medication "combination" arm (nicotine
patch plus bupropion plus a nicotine lozenge) or to a patch-only arm.
The primary endpoint is time to relapse (`ttr`/`relapse`); baseline
covariates include age, smoking history, prior cessation attempts,
gender, race, employment status, and a smoking intensity score.

```{r, message = FALSE}
data(pharmacoSmoking, package = 'asaur')

asaur_pharmacoSmoking <-
  pated(
    coxph_(Surv(time = ttr, event = relapse) ~ grp,           data_index = 1),
    glm_(age            ~ grp, family = "gaussian",           data_index = 1),
    glm_(yearsSmoking   ~ grp, family = "gaussian",           data_index = 1),
    glm_(priorAttempts  ~ grp, family = "gaussian",           data_index = 1),
    glm_(longestNoSmoke ~ grp, family = "gaussian",           data_index = 1),
    glm_(gender         ~ grp, family = "binomial",           data_index = 1),
    glm_(I(race == 'black')        ~ grp, family = "binomial", data_index = 1),
    glm_(I(race == 'hispanic')     ~ grp, family = "binomial", data_index = 1),
    glm_(I(race == 'white')        ~ grp, family = "binomial", data_index = 1),
    glm_(I(employment == 'ft')     ~ grp, family = "binomial", data_index = 1),
    glm_(I(employment == 'pt')     ~ grp, family = "binomial", data_index = 1),
    glm_(I(levelSmoking == 'heavy')~ grp, family = "binomial", data_index = 1),
    data = list(add_pid(pharmacoSmoking %>%
                          mutate(grp = ifelse(grp == 'combination', 1, 0))))
  )

printObject(asaur_pharmacoSmoking)
plot(asaur_pharmacoSmoking)
```

## Malignant glioma: `coin::glioma`

Grana et al. (2002) reported a small randomized trial of locoregional
radioimmunotherapy in patients with high-grade glioma (37 patients
total: 18 control, 19 radioimmunotherapy). Patients received standard
therapy with or without adjuvant radioimmunotherapy using `90`Y-biotin.
The endpoint is overall survival; available baseline covariates are
age, sex, and histology (glioblastoma multiforme vs. grade III
astrocytoma).

```{r, message = FALSE}
data(glioma, package = 'coin')

coin_glioma <-
  pated(
    coxph_(Surv(time = time, event = event)   ~ group, data_index = 1),
    glm_(age                ~ group, family = "gaussian", data_index = 1),
    glm_(sex                ~ group, family = "binomial", data_index = 1),
    glm_(I(histology == 'GBM') ~ group, family = "binomial", data_index = 1),
    data = list(add_pid(glioma %>%
                          mutate(group = ifelse(group == 'Control', 0, 1),
                                 event = 1 * event)))
  )

printObject(coin_glioma)
plot(coin_glioma)
```

## Burn-wound infection: `iBST::burn`

The burn data from Klein and Moeschberger (1997) come from a clinical
trial in which 154 severely burned patients (70 routine bathing,
84 chlorhexidine) were randomized to body cleansing with chlorhexidine
gluconate or to routine bathing care with soap. The primary endpoint
is time to staphylococcus infection (`T3`/`D3`). Baseline covariates
include sex (`Z2`), race (`Z3`), percentage of body surface burned
(`Z4`), location of the burn (head/buttock/trunk/upper-leg/lower-leg/
respiratory: `Z5`–`Z10`), and type of burn (`Z11`).

```{r, message = FALSE}
data(burn, package = 'iBST')

iBST_burn <-
  pated(
    coxph_(Surv(time = T3, event = D3) ~ Z1, data_index = 1),
    glm_(Z2 ~ Z1, family = "binomial", data_index = 1),
    glm_(Z3 ~ Z1, family = "binomial", data_index = 1),
    glm_(Z5 ~ Z1, family = "binomial", data_index = 1),
    glm_(Z6 ~ Z1, family = "binomial", data_index = 1),
    glm_(Z7 ~ Z1, family = "binomial", data_index = 1),
    glm_(Z8 ~ Z1, family = "binomial", data_index = 1),
    glm_(Z9 ~ Z1, family = "binomial", data_index = 1),
    glm_(Z10 ~ Z1, family = "binomial", data_index = 1),
    glm_(I(Z11 == 1) ~ Z1, family = "binomial", data_index = 1),
    glm_(I(Z11 == 2) ~ Z1, family = "binomial", data_index = 1),
    glm_(I(Z11 == 3) ~ Z1, family = "binomial", data_index = 1),
    glm_(Z4 ~ Z1, family = "gaussian", data_index = 1),
    data = list(add_pid(burn))
  )

printObject(iBST_burn)
plot(iBST_burn)
```

## Oropharyngeal carcinoma: `invGauss::d.oropha.rec`

The `d.oropha.rec` data describe time to recurrence in 192 patients
with oropharyngeal carcinoma (98 on regimen 1, 94 on regimen 2) who
were randomized between two treatment regimens (`treatm`). Available
baseline covariates are sex, age, T-stage, and N-stage.

```{r, message = FALSE}
data(d.oropha.rec, package = 'invGauss')

invGauss_d.oropha.rec <-
  pated(
    coxph_(Surv(time = time, event = status) ~ treatm, data_index = 1),
    glm_(I(sex == 1) ~ treatm, family = "gaussian", data_index = 1),
    glm_(age         ~ treatm, family = "gaussian", data_index = 1),
    glm_(tstage      ~ treatm, family = "gaussian", data_index = 1),
    glm_(nstage      ~ treatm, family = "gaussian", data_index = 1),
    data = list(add_pid(d.oropha.rec %>%
                          mutate(treatm = ifelse(treatm == 2, 1, 0))))
  )

printObject(invGauss_d.oropha.rec)
plot(invGauss_d.oropha.rec)
```

## ddI vs. ddC in advanced HIV: `JM::aids.id`

Abrams et al. (1994, *NEJM*) randomized HIV-infected patients who had
failed or were intolerant to zidovudine (AZT) to didanosine (ddI) or
zalcitabine (ddC). `aids.id` is the patient-level (one row per
subject) form distributed with the `JM` package and contains 467
subjects (237 ddC, 230 ddI). The endpoint is time to death, and the
baseline covariates used here are CD4 count, sex, prior opportunistic
infection status, and AZT history (failure vs. intolerance).

```{r, message = FALSE}
data(aids.id, package = 'JM')

JM_aids.id <-
  pated(
    coxph_(Surv(time = Time, event = death) ~ drug, data_index = 1),
    glm_(CD4    ~ drug, family = "gaussian", data_index = 1),
    glm_(gender ~ drug, family = "binomial", data_index = 1),
    glm_(I(prevOI == 'AIDS')        ~ drug, family = "binomial", data_index = 1),
    glm_(I(AZT  == 'intolerance')   ~ drug, family = "binomial", data_index = 1),
    data = list(add_pid(aids.id %>%
                          mutate(drug = ifelse(drug == 'ddC', 1, 0))))
  )

printObject(JM_aids.id)
plot(JM_aids.id)
```

## ACTG 320 antiretroviral trial: `multipleOutcomes::actg`

The `actg` dataset, redistributed with `multipleOutcomes` from the
`mlr3proba` package, corresponds to the ACTG 320 trial (Hammer et al.,
1997, *NEJM*), in which patients with CD4 counts below 200 were
randomized to a three-drug regimen (zidovudine + lamivudine +
indinavir) or to a two-drug control. The redistributed file contains
1151 patients (577 control, 574 three-drug). The composite endpoint
here is time to an AIDS-defining event or death, derived from the
`censor` and `censor_d` indicators. Baseline covariates include
stratification factor, sex, IV-drug history, race/ethnicity,
hemophilia, Karnofsky score, CD4 count, prior ZDV exposure, and age.

```{r, message = FALSE}
data(actg, package = 'multipleOutcomes')

mlr3proba_actg <-
  pated(
    coxph_(Surv(time = time, event = event) ~ tx, data_index = 1),
    glm_(strat2 ~ tx, family = "binomial", data_index = 1),
    glm_(sex    ~ tx, family = "binomial", data_index = 1),
    glm_(I(ivdrug == 1) ~ tx, family = "binomial", data_index = 1),
    glm_(I(raceth == 1) ~ tx, family = "binomial", data_index = 1),
    glm_(I(raceth == 2) ~ tx, family = "binomial", data_index = 1),
    glm_(I(raceth == 3) ~ tx, family = "binomial", data_index = 1),
    glm_(hemophil       ~ tx, family = "binomial", data_index = 1),
    glm_(I(karnof == 100) ~ tx, family = "binomial", data_index = 1),
    glm_(I(karnof == 90)  ~ tx, family = "binomial", data_index = 1),
    glm_(I(karnof == 80)  ~ tx, family = "binomial", data_index = 1),
    glm_(I(karnof == 70)  ~ tx, family = "binomial", data_index = 1),
    glm_(cd4      ~ tx, family = "gaussian", data_index = 1),
    glm_(priorzdv ~ tx, family = "gaussian", data_index = 1),
    glm_(age      ~ tx, family = "gaussian", data_index = 1),
    data = list(add_pid(actg %>%
                          mutate(event = 1 * (censor + censor_d > 0))))
  )

printObject(mlr3proba_actg)
plot(mlr3proba_actg)
```

## PBC3 cyclosporin A trial: `pec::Pbc3`

The PBC3 trial was a European multicentre randomized study (1983–1987)
comparing cyclosporin A (CyA) with placebo in 349 patients with
primary biliary cirrhosis (173 placebo, 176 CyA). The endpoint
analysed here is time to death or liver transplantation. Baseline
covariates include sex, histological stage, history of gastrointestinal
bleeding, age, and the biochemistry panel (creatinine, bilirubin,
alkaline phosphatase, aspartate transaminase, weight).

```{r, message = FALSE}
data(Pbc3, package = 'pec')

pec_Pbc3 <-
  pated(
    coxph_(Surv(time = days, event = event) ~ tment, data_index = 1),
    glm_(sex                 ~ tment, family = "binomial", data_index = 1),
    glm_(I(stage == 1)       ~ tment, family = "binomial", data_index = 1),
    glm_(I(stage == 2)       ~ tment, family = "binomial", data_index = 1),
    glm_(I(stage == 3)       ~ tment, family = "binomial", data_index = 1),
    glm_(I(stage == 4)       ~ tment, family = "binomial", data_index = 1),
    glm_(gibleed             ~ tment, family = "binomial", data_index = 1),
    glm_(age                 ~ tment, family = "gaussian", data_index = 1),
    glm_(crea                ~ tment, family = "gaussian", data_index = 1),
    glm_(bili                ~ tment, family = "gaussian", data_index = 1),
    glm_(alkph               ~ tment, family = "gaussian", data_index = 1),
    glm_(asptr               ~ tment, family = "gaussian", data_index = 1),
    glm_(weight              ~ tment, family = "gaussian", data_index = 1),
    data = list(add_pid(Pbc3 %>%
                          mutate(event = ifelse(status == 0, 0, 1))))
  )

printObject(pec_Pbc3)
plot(pec_Pbc3)
```

## GBSG-2 adjuvant breast-cancer trial: `pec::GBSG2`

The German Breast Cancer Study Group 2 trial (Schumacher et al.,
1994) randomized 686 women with node-positive primary breast cancer
(440 without hormonal therapy, 246 with tamoxifen) in a 2x2 factorial
of adjuvant hormonal therapy (tamoxifen, `horTh`) and chemotherapy
duration. The endpoint is recurrence-free survival; baseline
covariates include tumour size, number of positive lymph nodes,
progesterone receptor level, and tumour grade.

```{r, message = FALSE}
data(GBSG2, package = 'pec')

pec_GBSG2 <-
  pated(
    coxph_(Surv(time = time, event = cens) ~ horTh, data_index = 1),
    glm_(tsize    ~ horTh, family = "gaussian", data_index = 1),
    glm_(pnodes   ~ horTh, family = "gaussian", data_index = 1),
    glm_(progrec  ~ horTh, family = "gaussian", data_index = 1),
    glm_(I(tgrade == 'I')   ~ horTh, family = "binomial", data_index = 1),
    glm_(I(tgrade == 'II')  ~ horTh, family = "binomial", data_index = 1),
    glm_(I(tgrade == 'III') ~ horTh, family = "binomial", data_index = 1),
    data = list(add_pid(GBSG2 %>%
                          mutate(horTh = ifelse(horTh == 'yes', 1, 0))))
  )

printObject(pec_GBSG2)
plot(pec_GBSG2)
```

## Interpreting the output

For each trial, the printed table shows the adjusted PATED estimate
and the unadjusted Standard estimate of the treatment effect on the
primary endpoint, each with its standard error and two-sided p-value,
followed by one row per prognostic covariate. The `relativeEfficiency()`
helper turns the two standard errors into a single RE summary, and the
`plot()` method converts every row into a Wald z-statistic for a
visual diagnostic.

Read the two views together: the spread of the gray prognostic
z-values shows how well randomization balanced the covariates, while
the gap between the blue (unadjusted) and red (PATED) points
illustrates the variance reduction that the RE value quantifies. RE
close to 1 means the baseline covariates carried little prognostic
information beyond what randomization already balanced; RE
appreciably above 1 indicates a meaningful efficiency gain.

## References

- Abrams DI, Goldman AI, Launer C, et al. (1994). A comparative trial
  of didanosine or zalcitabine after treatment with zidovudine in
  patients with human immunodeficiency virus infection. *NEJM* 330:
  657–662.
- Grana C, Chinol M, Robertson C, et al. (2002). Pretargeted adjuvant
  radioimmunotherapy with yttrium-90-biotin in malignant glioma
  patients: a pilot study. *British Journal of Cancer* 86: 207–212.
- Hammer SM, Squires KE, Hughes MD, et al. (1997). A controlled trial
  of two nucleoside analogues plus indinavir in persons with human
  immunodeficiency virus infection and CD4 cell counts of 200 per
  cubic millimeter or less. *NEJM* 337: 725–733.
- Klein JP, Moeschberger ML (1997). *Survival Analysis: Techniques
  for Censored and Truncated Data*. Springer.
- Schumacher M, Bastert G, Bojar H, et al. (1994). Randomized
  2x2 trial evaluating hormonal treatment and the duration of
  chemotherapy in node-positive breast cancer patients. *Journal of
  Clinical Oncology* 12: 2086–2093.
- Steinberg MB, Greenhaus S, Schmelzer AC, et al. (2009).
  Triple-combination pharmacotherapy for medically ill smokers.
  *Annals of Internal Medicine* 150: 447–454.
